--- 
title: "Week 2 Problem Sheet" 
author: "Akasha Strub"
date: "05 July 2023"
output: html_document 
---

## Import libraries

```{r, message = FALSE}
# Import libraries
library(rethinking)
library(dplyr)
library(ggplot2)
library(dagitty)

# Set seed
set.seed(0507)
```

## Problem 1
### Part 1

From the Howell1 dataset, consider only the people younger than 13 years old. 
Assume that age influences weight through two paths. 
First, age influences height, and height influences weight. 
Second, age directly influences weight through age related changes in muscle growth and body proportions.
Draw the DAG that represents these causal relationships. 

```{r}
# Draw DAG
dag <- dagitty( "dag {
Age -> Height
Height -> Weight
Age -> Weight
}")

coordinates(dag) <- list(
  x = c(Age = 0, Height = 2, Weight = 1) , 
  y = c(Age = 0, Height = 0, Weight = 1))

drawdag(dag)
```

### Part 2

Write a generative simulation that takes age as an input and simulates height and
weight, obeying the relationships in the DAG.

```{r}
# Import Howell1 dataset
data(Howell1)

# Filter to people younger than 13
df <- Howell1 %>% 
  filter(age < 13)

# Standardize data
df1 <- df %>% 
  mutate(age_std = standardize(age),
         weight_std = standardize(weight),
         height_std = standardize(height))

# Find association between age and height to generate height as function of age
model1a <- quap(
  alist(
    height_std ~ dnorm(mu, sigma),
    mu <- a + b*age_std,
    a ~ dnorm( 0 , 1 ),
    b ~ dnorm( 0 , 1 ),
    sigma ~ dexp( 1 )
    ),
  data = df1)

# These posterior distribution estimates will be used for the generative model
# to generate height from age
precis(model1a)

# Find association between weight and age & height to generate weight as a function of age and height
model1b <- quap(
  alist(
    weight_std ~ dnorm(mu, sigma),
    mu <- a + b_age*age_std + b_height*height_std,
    a ~ dnorm( 0 , 1 ),
    b_age ~ dnorm( 0 , 1 ),
    b_height ~ dnorm( 0 , 1 ),
    sigma ~ dexp( 1 )
    ),
  data = df1)

# These posterior distribution estimates will be used for the generative model
# to generate weight from age and height
precis(model1b)

# Instantiate range of age values
age_std1 <- seq(-2, 2, length.out = 100)

# Model height as as function of age
height_std1 <- rnorm(0 + 0.92 * age_std1, 0.39)

# Model weight as a function of age and height
weight_std1 <- rnorm(0 + 0.07 * age_std1 + 0.90 * height_std1, 0.26)
```
### Part 3 

Estimate the causal association between age and weight.

```{r}
# For the direct causal association, height should be included to find the direct effect 
# of age on weight, for a fixed height. A multivariate regression is to be run (Gaussian 
# approximation of posterior will be used, as opposed to grid/MCMC)

# Create model for weight as a function on age
model2 <- quap(
  alist(
    weight_std ~ dnorm(mu, sigma),
    mu <- a + b_age*age_std + b_height*height_std,
    a ~ dnorm( 0 , 1 ),
    b_age ~ dnorm( 0 , 1 ),
    b_height ~ dnorm( 0 , 1 ),
    sigma ~ dexp( 1 )
    ),
  data = df1)

# Investigate validity of priors via prior predictive simulation
# Method 1 - plot relationships as lines for weight vs age and weight vs height
n_priors2 <- 100
prior_samples2a <- extract.prior(model2, n = n_priors2) %>%
  as.data.frame() %>%
  mutate(line = row_number())

df_plot2a <- expand.grid(line = seq(1:n_priors2),
                        age_std = seq(-2, 2)) %>%
  left_join(prior_samples2a, by = "line") %>%
  mutate(weight_std = a + b_age*age_std) %>%
  arrange(line, age_std)

# Plot
ggplot(data = df_plot2a, aes(x = age_std, y = weight_std, group = line)) +
  geom_line(alpha = 0.5) +
  labs(title = "Potential relationships between weight and age CONDITIONAL on height (from prior)")

df_plot2b <- expand.grid(line = seq(1:n_priors2),
                        height_std = seq(-2, 2)) %>%
  left_join(prior_samples2a, by = "line") %>%
  mutate(weight_std = a + b_height*height_std) %>%
  arrange(line, height_std)

# Plot
ggplot(data = df_plot2b, aes(x = height_std, y = weight_std, group = line)) +
  geom_line(alpha = 0.5) +
  labs(title = "Potential relationships between weight and height CONDITIONAL on age (from prior)")

# An additional prior predictive method for the multivariate regression use case 
# could be to simulate predictions from prior distributions and compare to raw data
# (this is less relevant for prior distributions as the points would be all over the place. 
# It is much more relevant for posterior distributions to evaluate model fit)

# See marginal posterior distributions
precis(model2)

# b_age is close to 0 and overlaps heavily with 0. Therefore CONDITIONAL on height,
# can can conclude that the DIRECT causal effect of age on height is minimal.
# This makes sense intuitively. If a child grows older but does not grow in height, 
# we would estimate that their weight would remain somewhat constant.
# And therefore that the direct causal effect be small. 
# The mean of 0.07 SD results in 0.1kg/year... "clinically insignificant" regardless
ggplot() + 
  geom_freqpoly(aes(x = extract.samples(model2, n = 10000)$b_age), stat = "bin") +
  geom_vline(aes(xintercept = 0), colour = "red") +
  labs(x = "b_age value",
       y = "freq",
       title = "Posterior distribution for b_age")
```

## Problem 2

Use a linear regression to estimate the total causal effect of each year of
growth on weight.

```{r}
# For the total causal association, Height should be ignored and a simple bivariate 
# regression is to be run (Gaussian approximation of posterior will be used, 
# as opposed to grid/MCMC)

# Create model for weight as a function on age
model3 <- quap(
  alist(
    weight_std ~ dnorm(mu, sigma),
    mu <- a + b*age_std,
    a ~ dnorm( 0 , 1 ),
    b ~ dnorm( 0 , 1 ),
    sigma ~ dexp( 1 )
    ),
  data = df1)

# Investigate validity of priors via prior predictive simulation
n_priors3 <- 100
prior_samples3 <- extract.prior(model3, n = n_priors3) %>% 
  as.data.frame() %>% 
  mutate(line = row_number())

df_plot3 <- expand.grid(line = seq(1:n_priors3),
                       age_std = seq(-2, 2), 
                       weight_std = NA) %>% 
  left_join(prior_samples3, by = "line") %>% 
  mutate(weight_std = a + b*age_std) %>% 
  arrange(line, age_std)

# Plot
ggplot(data = df_plot3, aes(x = age_std, y = weight_std, group = line)) + 
  geom_line(alpha = 0.5)

# See marginal posterior distributions
precis(model3)

# For every 1 SD increase in age (3.8 years), there is ~ 0.9 SD increase in weight (5.1kg). 
# The positive relationship is expected. Unstandardised, this results to ~ 1.34 kg per year
# df$age %>% sd()
# df$weight %>% sd()

```

## Problem 3

Now suppose the causal association between age and weight might be different for boys and girls. 
Use a single linear regression, with a categorical variable for sex, to estimate the total causal effect of age on weight separately for boys and girls. 
How do girls and boys differ? Provide one or more posterior contrasts as a summary.


```{r}
# For the total causal association, Height should again be ignored and a simple bivariate 
# regression is to be run. In this case, the parameter b should be specific to 
# the sex via an index variable (not dummy for various reasons...)

# Modify data for model
df2 <- df1 %>% 
  mutate(sex = case_when(male == 0 ~ 1,
                         male == 1 ~ 2))
         
# Create model for weight as a function on age
model4 <- quap(
  alist(
    weight_std ~ dnorm(mu, sigma),
    mu <- a[sex] + b[sex]*age_std,
    a[sex] ~ dnorm( 0 , 1 ),
    b[sex] ~ dnorm( 0 , 1 ),
    sigma ~ dexp( 1 )
    ),
  data = df2)

# Prior predictive simulation is redundant as it is the same as above - just for
# each gender individually this time around.

# See marginal posterior distributions
precis(model4, depth = 2)

# There seems to be a difference between male and female after all. Instead of
# only assessing MARGINAL distributions to quantify this, the best is to sample from 
# each posterior distribution and subtract to obtain a contrasts

# Create contrast distribution
posterior_contrast <- extract.samples(model4) %>% 
  as.data.frame() %>% 
  # Summarise as men coef - women coef for positive values (more intuitive)
  reframe(b_diff = b.2 - b.1)

# Plot contrast distribution between parameterss
ggplot() + 
  geom_freqpoly(aes(x = posterior_contrast$b_diff)) +
  geom_vline(aes(xintercept = 0), colour = "red") +
  labs(x = "b_male - b_female value",
       y = "freq",
       title = "Posterior distribution for difference in b values between M and F")

# Plot age vs weight lines
n_samples4 <- 1000
post_samples4 <- extract.samples(model4, n = n_samples4) %>% 
  as.data.frame() %>% 
  mutate(sample = row_number())

df_plot3 <- expand.grid(sex = c(1, 2),
                        age_std = seq(-1.41225, 1.725287),
                        sample = seq(1:n_samples4)) %>% 
  left_join(post_samples4, by = "sample") %>% 
  mutate(mean_weight_std = case_when(sex == 1 ~ a.1 + b.1*age_std,
                                     sex == 2 ~ a.2 + b.2*age_std)) %>%
  group_by(age_std, sex) %>% 
  reframe(mean_weight = mean(mean_weight_std),
          # Computing percentile intervals is just a matter of finding 5.5 and 94.5 
          # percentiles and then subtracting - rethinking package can do such simple
          # calculations on our behalf
          mean_weight_lower = PI(mean_weight_std)[1],
          mean_weight_upper = PI(mean_weight_std)[2]) %>% 
  arrange(age_std)

# Modify values back to real units
df_plot4 <- df_plot3 %>% 
  mutate(age = (age_std * df2$age %>% sd()) + df2$age %>% mean(),
         across(contains("weight"), ~ .x * df2$weight %>% sd()) + df2$weight %>% mean())

# Plot contrast over age range
ggplot(data = df_plot4, aes(x = age, y = mean_weight, group = sex)) +
  geom_line(aes(colour = factor(sex))) +
  geom_ribbon(aes(ymin = mean_weight_lower, ymax = mean_weight_upper, fill = factor(sex)), alpha = 0.2) +
  labs(title = "Non-standardised weight vs age for M vs F")
```
