--- 
title: "Week 1 Problem Sheet" 
author: "Akasha Strub"
date: "20 June 2023"
output: html_document 
---

## Import libraries

```{r, message = FALSE}
# Import libraries
library(rethinking)
library(dplyr)
library(ggplot2)
```

## Problem 1

Suppose the globe tossing data (Lecture 2, Chapter 2) had turned out to be 4 water and 11 land. Construct the posterior distribution.

```{r}
# Instantiate parameter as a grid
p_grid <- seq(from = 0, to = 1, length.out = 100)

# Define prior as uniform
prior <- rep(1, 100)

# Count number of ways data could have come about for each value of p_grid
# N.B. this could have been derived from first principles (garden of forking data)
# but in this case binomial distribution was used.
likelihood <- dbinom(4, size = 15, prob = p_grid)

# Compute unstandardised posterior (product of prior and likelihood)
unstd_posterior <- prior * likelihood

# Compute standardised posterior
posterior <- unstd_posterior / sum(unstd_posterior)

# Plot posterior
df <- data.frame(p_grid, posterior)
ggplot(data = df) + 
  geom_line(aes(x = p_grid, y = posterior)) +
  labs(title = "Posterior distribution",
       x = "Proportion of globe covered in water",
       y = "Density")
```

## Problem 2

Using the posterior distribution from 1, compute the posterior predictive distribution for the next 5 tosses of the same globe. I recommend you use the sampling method.

```{r}
# Sample 100 parameter values from posterior
post_samples <- sample(p_grid, size = 100, replace = TRUE, prob = posterior)

# For each posterior sample, use the binomial to create an observation
obs_samples <- rbinom(n = 1000, size = 5, prob = post_samples)

# Plot histogram
ggplot() + 
  geom_histogram(aes(obs_samples), binwidth = 0.5) +
  labs(title = "Posterior predictive distribution",
       x = "Number of 'W' throws in next 5 tosses",
       y = "Count")
```

## Problem 3

Use the posterior predictive distribution from 2 to calculate the probability of 3 or more water samples in the next 5 tosses.

```{r}
# Find proportion of predictive samples that result in 3 or more water tosses in next 5 tosses:
100 * sum(obs_samples >= 3) / length(obs_samples)

# Around 18.9% (depends on specific samples!)
```
